---
title: "p8105 Homework 2"
author: "sr4354"
date: "2026-10-05"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)

knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1
```{r import-transit}
nyc_transit <- read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")
```

```{r}
glimpse(nyc_transit)
```

```{r clean-transit}
nyc_transit <- nyc_transit |>
  select(
    Line,
    `Station Name`,
    `Station Latitude`,
    `Station Longitude`,
    starts_with("Route"),
    Entry,
    Vending,
    `Entrance Type`,
    ADA
  )
```

```{r}
glimpse(nyc_transit)
```

```{r clean-entry}
nyc_transit <- nyc_transit |>
  mutate(
    Entry = Entry == "YES"
  )
```

```{r}
glimpse(nyc_transit)
```

The NYC Transit dataset contains information on subway station entrances and exits, including the subway line, station name and location, routes served, entrance type, entry access, vending availability, and ADA accessibility. The original dataset contained 1,868 observations and 32 variables. I retained the variables relevant to the questions of interest and converted the Entry variable from a character variable indicating “YES” or “NO” to a logical variable. The resulting dataset contains 1,868 observations and 19 variables. Each observation represents a station entrance or exit, with station-level information repeated across entrances and exits. The data are not fully tidy because routes served by a station are stored across multiple variables (Route1 through Route11) rather than in a single route variable, so the route information will need to be reformatted for further analysis. 

```{r}
nyc_transit |>
  distinct(Line, `Station Name`) |>
  nrow()
```

There are 465 distinct subway stations in the dataset, where a station is defined by the combination of subway line and station name.

```{r}
nyc_transit |>
  filter(ADA == TRUE) |>
  distinct(Line, `Station Name`) |>
  nrow()
```

There are 84 distinct ADA-compliant subway stations in the dataset.

```{r}
table(nyc_transit$Vending)
```

```{r}
nyc_transit |>
  filter(Vending == "NO") |>
  summarise(
    proportion = mean(Entry)
  )
```

Among station entrances and exits without vending machines, 37.7% allow entry.

```{r}
nyc_routes <- nyc_transit |>
  pivot_longer(
    cols = Route1:Route7,
    names_to = "route_number",
    values_to = "route_name"
  )
```

```{r}
glimpse(nyc_routes)
```

```{r}
nyc_routes <- nyc_transit |>
  pivot_longer(
    cols = Route1:Route7,
    names_to = "route_number",
    values_to = "route_name"
  ) |>
  mutate(
    route_number = parse_number(route_number)
  )
```

```{r}
glimpse(nyc_routes)
```

```{r}
nyc_routes <- nyc_routes |>
  filter(!is.na(route_name))
```

```{r}
glimpse(nyc_routes)
```

```{r}
nyc_routes |>
  filter(route_name == "A") |>
  distinct(Line, `Station Name`) |>
  nrow()
```

There are 60 distinct subway stations that serve the A train.

```{r}
nyc_routes |>
  filter(route_name == "A", ADA == TRUE) |>
  distinct(Line, `Station Name`) |>
  nrow()
```

Of the 60 stations serving the A train, 17 are ADA compliant.

```{r}
nyc_routes <- nyc_routes |>
  select(-Route8, -Route9, -Route10, -Route11)
```

```{r}
glimpse(nyc_routes)
```

## Problem 2

```{r}
excel_sheets("data/202610 Trash Wheel Collection Data.xlsx")
```

```{r}
mr_trash_wheel <- read_excel(
  "data/202610 Trash Wheel Collection Data.xlsx",
  sheet = "Mr. Trash Wheel"
)
```

```{r}
glimpse(mr_trash_wheel)
```

```{r}
mr_trash_wheel <- mr_trash_wheel |>
  select(1:14)
```

```{r}
glimpse(mr_trash_wheel)
```
```{r}
mr_trash_wheel <- mr_trash_wheel |>
  rename(
    dumpster = Dumpster,
    month = Month,
    year = Year,
    date = Date,
    weight_tons = `Weight (tons)`,
    volume_cubic_yards = `Volume (cubic yards)`,
    plastic_bottles = `Plastic Bottles`,
    polystyrene = Polystyrene,
    cigarette_butts = `Cigarette Butts`,
    glass_bottles = `Glass Bottles`,
    plastic_bags = `Plastic Bags`,
    wrappers = Wrappers,
    sports_balls = `Sports Balls`,
    homes_powered = `Homes Powered*`
  )
```

```{r}
glimpse(mr_trash_wheel)
```
```{r}
sum(is.na(mr_trash_wheel$dumpster))
```

```{r}
glimpse(mr_trash_wheel)
```


```{r}
mr_trash_wheel <- mr_trash_wheel |>
  filter(!is.na(dumpster))
```

```{r}
dim(mr_trash_wheel)
```

```{r}
mr_trash_wheel <- mr_trash_wheel |>
  mutate(
    sports_balls = as.integer(round(sports_balls))
  )
```

```{r}
glimpse(mr_trash_wheel)
```
```{r}
professor_trash_wheel <- read_excel(
  "data/202610 Trash Wheel Collection Data.xlsx",
  sheet = "Professor Trash Wheel"
)
```

```{r}
glimpse(professor_trash_wheel)
```

```{r}
professor_trash_wheel <- professor_trash_wheel |>
  rename(
    dumpster = Dumpster,
    month = Month,
    year = Year,
    date = Date,
    weight_tons = `Weight (tons)`,
    volume_cubic_yards = `Volume (cubic yards)`,
    plastic_bottles = `Plastic Bottles`,
    polystyrene = Polystyrene,
    cigarette_butts = `Cigarette Butts`,
    glass_bottles = `Glass Bottles`,
    plastic_bags = `Plastic Bags`,
    wrappers = Wrappers,
    homes_powered = `Homes Powered*`
  )
```

```{r}
glimpse(professor_trash_wheel)
```
```{r}
professor_trash_wheel <- professor_trash_wheel |>
  mutate(
    plastic_bottles = as.numeric(plastic_bottles)
  )
```


```{r}
professor_trash_wheel <- professor_trash_wheel |>
  filter(!is.na(dumpster))
```

```{r}
professor_trash_wheel <- professor_trash_wheel |>
  mutate(
    trash_wheel = "Professor"
  )
```

```{r}
glimpse(professor_trash_wheel)
```

```{r}
mr_trash_wheel <- mr_trash_wheel |>
  mutate(
    trash_wheel = "Mr"
  )
```

```{r}
gwynnda_trash_wheel <- read_excel(
  "data/202610 Trash Wheel Collection Data.xlsx",
  sheet = "Gwynnda the Good Wheel of the W"
)
```

```{r}
glimpse(gwynnda_trash_wheel)
```

```{r}
gwynnda_trash_wheel <- gwynnda_trash_wheel |>
  rename(
    dumpster = Dumpster,
    month = Month,
    year = Year,
    date = Date,
    weight_tons = `Weight (tons)`,
    volume_cubic_yards = `Volume (cubic yards)`,
    plastic_bottles = `Plastic Bottles`,
    polystyrene = Polystyrene,
    cigarette_butts = `Cigarette Butts`,
    plastic_bags = `Plastic Bags`,
    wrappers = Wrappers,
    homes_powered = `Homes Powered*`
  )
```

```{r}
glimpse(gwynnda_trash_wheel)
```
```{r}
sum(is.na(gwynnda_trash_wheel$dumpster))
```

```{r}
gwynnda_trash_wheel <- gwynnda_trash_wheel |>
  filter(!is.na(dumpster))
```

```{r}
gwynnda_trash_wheel <- gwynnda_trash_wheel |>
  mutate(
    trash_wheel = "Gwynnda"
  )
```

```{r}
glimpse(gwynnda_trash_wheel)
```

```{r}
mr_trash_wheel <- mr_trash_wheel |>
  mutate(
    year = as.numeric(year)
  )
```

```{r}
trash_wheels <- bind_rows(
  mr_trash_wheel,
  professor_trash_wheel,
  gwynnda_trash_wheel
)
```

```{r}
glimpse(trash_wheels)
```



